---
title: "Peer Assessment 1"
author: "-Pareen Jain"
date: "17 July 2015"
output: html_document
---
###Loading and preprocessing the data
- Load the data
The original data set:
```{r}
library(knitr)
if(!file.exists("PA1")){ dir.create("PA1")}
setwd("E:/CourseraDataScience/Workspace/PA1")
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileurl, destfile = "E:/CourseraDataScience/Workspace/PA1/activity.zip", method = "curl")
activity <- read.csv(unz("activity.zip", "activity.csv"), header=TRUE)
str(activity)
```

- Process/transform the data into a suitable format.
We change the class of the date variable from factor into 'Date'.
The processed data set:
```{r}
activity$date <- as.Date(as.character(activity$date))
str(activity)
```

###What is mean total number of steps taken per day?
-Calculate the total number of steps taken per day
```{r}
steps_day <- tapply(activity$steps, activity$date, sum)
```

-Make a histogram of the total number of steps taken each day
```{r}
hist(steps_day, main = "Histogram of Total Number of Steps taken each day", xlab = "Total Number of Steps", col = "turquoise")
```

-Calculate and report the mean and median of the total number of steps taken per day
```{r}
Mean <- mean(steps_day, na.rm = TRUE)
Median <- median(steps_day, na.rm = TRUE)
```
The mean is `r Mean` and the median is `r Median`.

###What is the average daily activity pattern?
-Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
steps_interval <- aggregate(steps ~ interval, activity, mean)
plot(steps_interval$interval, steps_interval$steps, type = "l", lwd = 1, main = "Average Number of Steps Averaged across all days", xlab = "Interval", ylab = "Average Number of Steps")
```

-Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_steps <- steps_interval[which.max(steps_interval$steps), ]["interval"]
```
The interval with the maximum number of steps is the `r max_steps`th interval.

###Imputing missing values
-Calculate and report the total number of missing values in the dataset
```{r}
NAs <- sum(!complete.cases(activity))
```
There are `r NAs` rows with missing values in the data set.

-Filling in all of the missing values in the dataset using the mean for that 5-minute interval and Creating a new Dataset.
```{r}
activity_new <- activity
for(i in 1:nrow(activity_new)) {
  if(is.na(activity_new$steps[i])) {
    activity_new$steps[i] <- steps_interval$steps[which(steps_interval$interval == activity_new$interval[i])]
  }
}
```

-Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r}
steps_day_new <- tapply(activity_new$steps, activity_new$date, sum)
hist(steps_day_new, main = "New Histogram of Total Number of Steps taken each day", xlab = "Total Number of Steps", col = "turquoise")
Mean_new <- mean(steps_day_new, na.rm = TRUE)
Median_new <- median(steps_day_new, na.rm = TRUE)
```
The new mean is `r Mean_new` and the new median is `r Median_new`.

###Are there differences in activity patterns between weekdays and weekends?
-Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
activity_new$day <- weekdays(activity_new$date)
activity_new$daytype <- c()
for(i in 1:nrow(activity_new)) {
  if(activity_new$day[i] == "Saturday" | activity_new$day[i] == "Sunday") {
    activity_new$daytype[i] <- "weekend"
  }
  else{
    activity_new$daytype[i] <- "weekday"
  }
}
activity_new$daytype <- as.factor(activity_new$daytype)
```

-Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
par(mfrow = c(2, 1))
steps_interval_new <- aggregate(steps ~ interval + daytype, data = activity_new, mean)
weekend_set <- subset(steps_interval_new, daytype == "weekend")
weekday_set <- subset(steps_interval_new, daytype == "weekday")
plot(weekday_set$interval, weekday_set$steps, type = "l", lwd = 1, main = "Average Number of Steps Averaged across Week days", xlab = "Interval", ylab = "Average Number of Steps", xlim = c(0, 2500), ylim = c(0, 250))
plot(weekend_set$interval, weekend_set$steps, type = "l", lwd = 1, main = "Average Number of Steps Averaged across Weekends", xlab = "Interval", ylab = "Average Number of Steps", xlim = c(0, 2500), ylim = c(0, 175))

```

